Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-30 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Completed JED revisions Ch 9 -11

The goal of this chapter is to utilize a numerical tool to predict or simulate the results from some of the key experiments conducted in this study.  Computational fluid dynamics (CFD) models are not as widely used by fire investigators as the algebraic tools are.  The CFD models require more knowledge, more computer resources, and more input data to utilize correctly.  Currently modelers need to develop their own data or rely on loosely defined databases with the hope that the material tested in the database is similar to the specific material of interest.

@@

The Fire Dynamics Simulator (FDS) and the scientific visualization tool, Smokeview, are the most widely used and well documented CFD tools in the fire protection engineering community. In addition, these tools have on going verification and validation programs.  Given the extent and high quality of the existing FDS documentation, there is no value to attempting to recreate a description of the model here other than to direct the reader to the volumes of information and data repository that exist for the FDS and Smokeview suite.

FDS~\cite{FDS_Users_Guide} is a CFD model initially developed and currently maintained by NIST that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires. A CFD model has a computational domain that is divided into  three-dimensional rectangular control volumes or computational cells. The size and arrangement of computational cells is sometimes referred to as the grid or the mesh.  The CFD model computes the density, velocity, temperature, pressure, and gas concentrations in each cell. Based on the laws of conservation of mass, momentum, and energy, the model tracks the generation and movement of fire gases. One of the most important aspects of FDS is that it is mathematically verified~\cite{FDS_Verification_Guide}and validated against fire test data~\cite{FDS_Validation_Guide}to ensure that it is accurate and provides the expected results, given appropriate input data.  The data that FDS has been validated against can be found on-line is repository of the reports that were the source of the data~\cite{FDS-SMV_repository}. A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Tech_Guide}.

Smokeview is a software tool specifically designed to visualize simulation results from FDS~\cite{Smokeview_Users_Guide}. Smokeview visualizes smoke and other attributes of the fire simulation by showing tracer particle flow and two dimensional (2D) or three dimensional (3D) shaded contours of gas flow data, including temperature and flow vectors showing flow direction and magnitude. Smokeview has the ability to visualize fire and smoke by displaying a series of partially transparent planes where the transparencies in each plane (at each grid node) are determined from soot densities computed by FDS. Smokeview can also visualize data at particular snapshots in time using 2D or 3D contours of data that show temperature, flow direction, and flow magnitude.  Smokeview also has a feature that can represent a fire or char pattern on a surface based on user supplied threshold criteria.  Smokeview also has a Technical Reference Guide~\cite{Smokeview_Tech_Guide} and a Verification Guide~\cite{Smokeview_Verification_Guide}.

@@

One of the key determinations that is required when developing an FDS model is the mesh or grid cell size.  If a single plane of the data from the model were compared to a digital image we might refer to the grid cell size as the resolution of the image.  The smaller the grid cells the better the resolution. However increased resolution will require more computational resources and additional run time. The question is, does it provide a more realistic answer?  As with most pursuits there is no one answer.  The `best' grid size may depend on a several different parameters of the scenario being modeled.

Choosing the size of the grid when performing an analysis with FDS is critical in order to get the level of resolution needed to adequately simulate the required details within a simulation. In studies where large rooms or portions of buildings have been modeled grid cells on the order of 100~mm have been used and resulted in the successful simulation of fire flow and fire growth trends ~\cite{Grosshandler:Station,Madrzykowski:2004,Vettori:2000, Barowy:2012,Overholt:San_Francisco,Weinschenk:Chicago}.

@@

To examine the impact of computational grid cell size on the results, three different computational cell sizes; 80~mm, 40~mm, and 20~mm were chosen to model the fire. The resulting resolution index (RI) values were approximately 4.4, 8.8, and 17.5, respectively. Given the range of grid cell sizes and the characteristic diameter of fire, 0.35~m, this resulted in what could be considered a course (RI=5), medium (RI=10), and fine (RI=20) resolution.  FDS simulations of the source fire characterization experiments will be conducted with these resolutions and compared with the results from the experiments.

@@

The FDS model for simulating the source fires is very basic, consisting of the definition of the computational domain in space and time, a definition of the burner geometry, the combustion model for the fuel, and the specification of measurement points that can be used to examine temperature, heat flux, and flame height.

The computational domain of the model had the geometry of a cube with 2.4~m on a side.  The resulting number of computational cells for each case ranged from 27,000 for the 80~mm cells to 1,728,000 for the 20~mm cells. For the 80 and 40~mm computational cell scenarios, the cases were run as a single mesh.  For the 20~mm case, the computational domain was divided into 8 uniform meshes.  Each mesh was 2.4~m by 1.2~m and 0.6~m tall.  The time domain for each model was chosen to be 300 seconds, which is similar to the burn time for each of the fuels.

The burner was modeled as an obstruction with a open top area of 0.09~m$^2$, elevated 0.08~m above the test surface.  This is the same physical size of the natural gas burner used in the experiments.  It was modeled with a steady state heat release rate flux of 888.89 kW/m$^2$, or a heat release rate of 80~kW.   The chosen fuel reaction was `methane' since the natural gas used in the laboratory experiments contained approximately 95\% methane.  In FDS version 6, the default radiant fraction for the `methane' reaction was set to~0.2.

FDS has devices, which are used to simulate thermocouples and total heat flux gauges.  These devices were positioned in locations similar to the locations used in the source fire experiments.  The thermocouples were positioned in a vertical array, centered on the burner, beginning 0.1~m above the burner surface and spaced every 0.1~m up to 1.2~m above the burner surface.  A total heat flux gauge was also modeled to provide data at the 0.95 m from the centerline of the burner and 0.5 m above the top of the burner.  
In addition, FDS features devices that can not be physically implemented.  One such device, is used to determine the heat release rate per unit length.  This device was used to determine flame height~\cite{McDermott:2011}.

@@

In Chapter 2 the fires from the burners were characterized in terms of plume temperature, heat flux remote from the plume, and flame height.  In this section the results from FDS simulations of the fire generated by the natural gas burner experiments are compared with the experimental results.  The average plume temperatures at elevations from 0.1~m through 1.2~m above the top face of the burner in 0.1~m increments will be compared.

\subsection{Grid Size Analysis}
Figure~\ref{FHNG01_RI} shows the comparison of the centerline gas burner temperatures with three different sets of FDS results.  The only variable between these simulations was the RI. The RIs included three different computational cell sizes; 80~mm (RI=5), 40~mm~(RI=10), 20~mm~(RI=20).  The associated 95\% confidence limits are also shown the temperature value for each height.  The FDS simulated temperatures based on the course cell size of 80~mm (RI=5) had the best agreement with the temperatures along centerline of the plume.  The two simulations with the higher RI values provided less agreement than the course mesh.

@@

This seemed to be counter intuitive to the results expected from an increased RI.  However in each of the FDS models, the results closest to the burner seem to indicate that the model may not be `burning' the appropriate amount of fuel (fuel rich), which resulted in temperatures significantly lower than measured.  The effect of this fuel rich area above the burner results in poorer agreement with increasing RI.  It appears that the the RI=10 and RI=20 models follow the correct trend, however the combustion zone was moved to a higher elevation relative to what was measured.  For example the values at 0.5~m and 0.8~m above the burner, from the RI=10 and RI=20 cases respectively, seemed to coincide with the measured value at 0.3~m above the burner.  Adjusting the FDS corresponding predictions from the RI=10 and RI=20 models down to the 0.3~m position appears to move the remaining temperatures above those points into alignment with the measurements along the centerline of the plume. As FDS is a complex model, this explanation may be only one possibility.

These simulation results are consistent with the comparisons made to the McCaffrey's plume correlation from his study, `Purely Buoyant Diffusion Flames'~\cite{McCaffrey:1979}, in the FDS Validation Guide~\cite{FDS_Validation_Guide}. McCaffrey used a natural gas burner, 0.30~m square, which is similar in size to the burner used in this study.  McCaffrey noted that, "visually non-axisymmetric behavior could be observed only very close to the burner surface".  In the McCaffrey study, temperatures were measured with Type K thermocouples covered with a 1.0~mm stainless steel sheath. The peak heat release rate in McCaffery's plume study was 57.5~kW, this heat release rate was also the closest to the 80~kW fires in the current study.  The comparison between the McCaffrey plume correlation and the FDS simulations at RI values of 5, 10 and 20 are shown in Figure~\ref{McCaffrey_Plume_comp}. Focusing on the data between the burner surface and the 1~m above the burner surface, FDS under predicted the temperature near the surface of the burner and then over predicted as the elevation increased to approximately 0.4~m and 1.3~m above the burner for the RI=10 and 20 cases respectively. Once the temperature comparisons were above the continuous and intermittent flame regions, the agreement between the cases and the correlation began to improve.  Similar to the current study, increased RIs did not improve the simulation of the vertical centerline temperatures close to the burner.

\begin{figure}[h]
	\includegraphics[width=5in] {../Figures/McCaffrey_Plume_comp}  \\
	\caption[Comparison of vertical centerline temperatures above a 0.3~m natural gas burner from McCaffrey with FDS simulated gas temperatures.] {Comparison of vertical centerline temperatures above a 0.3~m natural gas burner from McCaffrey with FDS simulated gas temperatures generated with three different RIs 5, 10, and 20 from the FDS Validation Guide~\cite{FDS_Validation_Guide}.}
	\label{McCaffrey_Plume_comp}
\end{figure}

The comparisons above used the predicted gas temperature from the model. Another comparison can be made examining the `THERMOCOUPLE' results from FDS to see if the model of the thermocouple provides improved agreement with the temperatures measured via physical thermocouples.  This comparison is shown in Figure~\ref{FHNG01_RI_TC}.  Not surprisingly, the sub model of the thermocouple device shifted the FDS values slightly.  At temperatures greater than 700$^\circ$C, the `THERMOCOUPLE' temperatures tended to be lower than the gas temperatures, perhaps due to radiant heat transfer losses. At temperatures less than 700$^\circ$C, the `THERMOCOUPLE' temperatures tended to be a higher than the gas temperatures, perhaps due to gains from radiant heat transfer.  The best agreement along the vertical array of temperatures was provided by the RI=5 model.  Variables such as density and emissivity required for modeling the thermal response and heat transfer to the thermocouple bead may be generating the differences between the gas temperature and the measured temperature FDS values.

\begin{figure}[h]
	\includegraphics[width=5in]{../Figures/FHNG01_RI_TC} \\
	\caption[Comparison of vertical centerline `THERMOCOUPLE' temperatures above the natural gas burner with FDS simulated gas temperatures.]{Comparison of vertical centerline `THERMOCOUPLE' temperatures above the natural gas burner with FDS simulated gas temperatures generated with three different RIs 5, 10, and 20.}
	\label{FHNG01_RI_TC}
\end{figure}

@@

The heat flux data being used for comparison was from the Chapter 2 source fire characterization experiments, the measured heat flux at 0.95~m from the centerline of the burner and 0.5~m above the top of the burner.  In Figure~\ref{FHNG01_HF_RI} the averaged values from both heat flux gages are plotted against the FDS calculated heat flux at three different RIs of 5, 10 and 20.  These FDS simulations used the default radiative fraction value of~0.2.  As with the temperature comparisons, the course grid with an RI=5 provided the best agreement to the measurements, however all of the values were lower than the measured heat flux.

\begin{figure}
	\includegraphics[width=5in]{../Figures/FHNG01_HF_RI} \\
	\caption[Comparison of the average heat flux measured 0.95~m from the centerline and 0.5~m above the top of the natural gas burner with FDS simulated heat flux values.]{Comparison of the average heat flux measured 0.95~m from the centerline and 0.5~m above the top of the natural gas burner with FDS simulated heat flux values generated with three different RIs 5, 10, and 20.}
	\label{FHNG01_HF_RI}
\end{figure}

Reviewing the heat flux data documented in Chapter 2 from both the 0.95~m  and the 2.0~m from the vertical centerline of the burner positions provided average steady state heat flux values of approximately 2.0~kW/m$^2$ and 0.43~kW/m$^2$, respectively.  Applying the point source method, discussed in Chapter 7, values for the radiative fraction were determined to be in the range of 0.25 to 0.28.  Values found in the literature for natural gas range from 0.15 to 0.25~\cite{Modak:1977,Hostikka:3}.   
Therefore it was of interest to conduct FDS simulations with radiative fractions ranging from 0.2 to 0.3.  This was based on the default value used in FDS and the upper bound of the literature values as this range covered the radiative fraction values estimated from the experiments.  The course grid, RI=5, was used for the FDS models.  The comparison of the results is shown in Figure~\ref{FHNG01_HF_RI5}.  The radiative fraction of 0.22 provided the best agreement with the measurement.

@@

One of the outputs that can be gotten from FDS is a heat release rate per unit length (HRRPUL), it has been used to determine the height where energy stops being released along the vertical centerline of the gas burner~\cite{McDermott:2011}.  Estimates of the flame height were made by estimating the height where 95 to 99~$\%$ of the total energy had been released. To have a value, based on a measurement, for comparison to the HRRPUL the peak visible flame height range estimate was developed.  The peak visible flame height ranged from the location where the flame existed only 10\% of the time with the 95\% certainty range.  The Table~\ref{tab:FDS_FH} displays estimated values from the models using all three RI values as well as the 95~$\%$ certainty range for the peak visible flame height from Chapter 2.  The course and medium grids, RI=5 and R=10, were not significantly different from the estimations of the peak visible flame height 95~$\%$ certainty value range.

@@

One of the advantages of a computational fluid dynamics model such as FDS is the ability produce enough spatial data in 3 dimensions to allow the generation of visualizations.  The impact of the mesh resolution can be seen in the three images of the fire shown in Figure~\ref{FDS_Flame_Image_comp}.  The increased resolution leads to a flame structure that provides a better match to the look of the actual flame.  However in this scenario, the increased resolution within the flow field has not resulted in improved numerical agreement with the measured values.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=1.7in]{../Figures/NG_80kW_GBWall_2D_RI=05_0601crop}
  \includegraphics[width=1.7in]{../Figures/NG_80kW_GBWall_2D_RI=10_0601crop}
  \includegraphics[width=1.7in]{../Figures/NG_80kW_GBWall_2D_RI=20_0601crop}\\
  \caption[Flame images generated with FDS and Smokeview]{Flame images generated with FDS and Smokeview. The left image has a grid size of 80~mm (RI=5), the center image has a grid size of 40~mm (RI=10) and the image on the right has a grid size of 20~mm (RI=20).}
  \label{FDS_Flame_Image_comp}
\end{figure}

@@

The comparisons of the temperature, heat flux, and flame heights for this specific scenario showed the best agreement with the 80~mm grid (RI=5) FDS simulations.  Even though the differences between the radiative fractions of 0.2 and 0.22 were small, for the remaining simulation cases, 0.22 will be used.  While the best agreement for the fire was demonstrated with the RI=5 grid, the RI=10 grid will also be used in the subsequent instrumented wall and the compartment model sections to see if the increased resolution comes into play as the models get additional obstructions and surfaces.

@@

The next comparisons were made with the natural gas source fire at 4 different distances from an instrumented gypsum board wall.  Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610~m), one burner width (0.305~m), one half burner width (0.150~m), and then touching the wall (0.0~m).  The simulations will replicate the same distances.  The simulation included a vertical gypsum board wall, 1.2~m wide, that extends from the floor of the model up to the top of the domain.  The material properties of the wall are based solely on the properties of gypsum that were cited in Chapter 4.  The values that are compared in the section include: vertical centerline temperatures above the burner, wall temperatures, and heat flux values at the position of the wall surface.

@@

The computational domain of the model was the same as the volume considered for the source fire simulations, a cube with 2.4~m on a side. The models were run with two different sized computational grid cells, an 80~mm (RI=5) grid and a 40~mm (RI=10) grid.  The face of the wall is located 1.95~m from the front of the domain.  
The wall was modeled with material properties for gypsum wallboard that were documented in Chapter 4.  The input properties were: thermal conductivity (0.15 W/m K), specific heat (1.00 kJ/kg K) and density (807 kg/m$^3$).  In addition FDS includes a default value of 0.9 for emissivity, this value should be representative of the painted surface of the gypsum wallboard. As noted earlier in this study, gypsum wallboard undergoes chemical and physical changes as it is heated.  These changes include the loss of homogeneity of the gypsum material, a change in density and as the paper is charred or burned there is a potential for a change in emissivity.  The thermal conductivity and the specific heat of the gypsum wallboard varies on the order of 20~$\%$ over the temperature range from 20 to 500$^\circ$C~\cite{Gross:1985}.  While FDS can accommodate changes in thermo-conductivity and specific heat with respect to time, density and emissivity must be held constant~\cite{FDS_Users_Guide}.  For the purposes of this study, the input values will remain constant throughout the course of the simulations.  The only variable will be the distance between the edge of the burner and the wall.

As before, the FDS simulations included devices which are used to simulate thermocouples and total heat flux gauges.  These devices were positioned in the same locations as the locations used in the physical experiments, above the centerline of the burner and on the wall as shown in Chapter 4, Figure~\ref{Instrumented_GB_Wall}. In addition, HRRPUL will again be used as a means to gain insight into the peak flame height simulated by FDS.

@@

 The temperatures measured by thermocouples positioned along the vertical centerline of the natural gas burner will be compared with FDS generated temperatures from two simulations, one with RI-5 and one with RI=10.  The measurement points were at intervals of 0.2~m from 0.2~m to 2.2~m above the burner.  Figure~\ref{FDS_Plume_IWGB_comp} contains four graphs.  Each graph shows the same types of comparison, the only difference is the distance away from the gypsum wallboard.  Beginning with the top left graph position and viewing left to right the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.

>>>>

The goal of this chapter is to utilize a numerical tool to predict or simulate the results from some of the key experiments conducted in this study.  Computational fluid dynamics (CFD) models are not as widely used by fire investigators as the algebraic tools are.  The CFD models require more knowledge, more computer resources, and more input data to utilize correctly.  Currently modelers must develop their own data or rely on loosely defined databases with the hope that the material tested in the database is similar to the specific material of interest.

@@

The Fire Dynamics Simulator (FDS) and the scientific visualization tool, Smokeview, are the most widely used and well-documented CFD tools in the fire protection engineering community. In addition, these tools have on-going verification and validation programs.  Given the extent and high quality of the existing FDS documentation, there is no value to attempting to recreate a description of the model here other than to direct the reader to the volumes of information and data repository that exist for the FDS and Smokeview suite.

FDS~\cite{FDS_Users_Guide} is a CFD model initially developed and currently maintained by NIST that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow, with an emphasis on smoke and heat transport from fires. A CFD model has a computational domain that is divided into  three-dimensional rectangular control volumes or computational cells. The size and arrangement of computational cells is sometimes referred to as the grid or the mesh.  The CFD model computes the density, velocity, temperature, pressure, and gas concentrations in each cell. Based on the laws of conservation of mass, momentum, and energy, the model tracks the generation and movement of fire gases. One of the most important aspects of FDS is that it is mathematically verified~\cite{FDS_Verification_Guide}and validated against fire test data~\cite{FDS_Validation_Guide} to ensure that it is accurate and provides the expected results, given appropriate input data.  The data FDS has been validated against can be found on-line in a repository of the reports that were the source of the data~\cite{FDS-SMV_repository}. A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Tech_Guide}.

Smokeview is a software tool specifically designed to visualize simulation results from FDS~\cite{Smokeview_Users_Guide}. Smokeview visualizes smoke and other attributes of the fire simulation by showing tracer particle flow and two dimensional (2D) or three dimensional (3D) shaded contours of gas flow data, including temperature and flow vectors showing flow direction and magnitude. Smokeview can visualize fire and smoke by displaying a series of partially transparent planes in which the transparencies in each plane (at each grid node) are determined from soot densities computed by FDS. Smokeview can also visualize data at particular snapshots in time using 2D or 3D contours of data that show temperature, flow direction, and flow magnitude.  Smokeview also has a feature that can represent a fire or char pattern on a surface based on user-supplied temperature threshold criteria.  Smokeview also has a Technical Reference Guide~\cite{Smokeview_Tech_Guide} and a Verification Guide~\cite{Smokeview_Verification_Guide}.

@@

One of the key determinations required when developing an FDS model is the mesh or grid cell size.  If a single plane of the data from the model were compared to a digital image, we might refer to the grid cell size as the resolution of the image.  The smaller the grid cells, the better the resolution. However, increased resolution will require more computational resources and additional run time. The question is, does it provide a more realistic answer?  As with most pursuits there is no one answer.  The best grid size may depend on a several different parameters of the scenario being modeled.

Choosing the size of the grid when performing an analysis with FDS is critical in order to get the level of resolution needed to adequately simulate the required details within a simulation. In studies in which large rooms or portions of buildings have been modeled, grid cells on the order of 100~mm have been used and resulted in the successful simulation of fire flow and fire growth trends ~\cite{Grosshandler:Station,Madrzykowski:2004,Vettori:2000, Barowy:2012,Overholt:San_Francisco,Weinschenk:Chicago}.

@@

To examine the impact of computational grid cell size on the results, three different computational cell sizes (80~mm, 40~mm, and 20~mm), were chosen to model the fire. The resulting resolution index (RI) values were approximately 4.4, 8.8, and 17.5, respectively. Given the range of grid cell sizes and the characteristic diameter of fire, 0.35~m, this resulted in what could be considered a course (RI=5), medium (RI=10), and fine (RI=20) resolution.  FDS simulations of the source fire characterization experiments were conducted with these resolutions and compared with the results from the experiments.

@@

The FDS model for simulating the source fires is very basic, consisting of the definition of the computational domain in space and time, a definition of the burner geometry, the combustion model for the fuel, and the specification of measurement points used to examine temperature, heat flux, and flame height.

The computational domain of the model had the geometry of a cube with 2.4~m on a side.  The resulting number of computational cells for each case ranged from 27,000 for the 80~mm cells to 1,728,000 for the 20~mm cells. For the 80 and 40~mm computational cell scenarios, the cases were run as a single mesh.  For the 20~mm case, the computational domain was divided into eight uniform meshes.  Each mesh was 2.4~m x 1.2~m, and 0.6~m tall.  The time domain for each model was chosen to be 300 seconds, which is similar to the burn time for each of the fuels.

The burner was modeled as an obstruction with a open top area of 0.09~m$^2$, elevated 0.08~m above the test surface.  This is the same physical size of the natural gas burner used in the experiments.  It was modeled with a steady state heat release rate flux of 888.89 kW/m$^2$, or a heat release rate of 80~kW.   The chosen fuel reaction was METHANE since the natural gas used in the laboratory experiments contained approximately 95\% methane.  In FDS version 6, the default radiant fraction for the METHANE reaction was set to~0.2.

FDS has devices used to simulate thermocouples and total heat flux gauges.  These devices were positioned in locations similar to the locations used in the source fire experiments.  The thermocouples were positioned in a vertical array, centered on the burner, beginning 0.1~m above the burner surface and spaced every 0.1~m up to 1.2~m above the burner surface.  A total heat flux gauge was also modeled to provide data at 0.95 m from the centerline of the burner and 0.5 m above the top of the burner.  
In addition, FDS features devices that can not be physically implemented.  One such device is used to determine the heat release rate per unit length.  This device was used to determine flame height~\cite{McDermott:2011}.

@@

In Chapter 2, the fires from the burners were characterized in terms of plume temperature, heat flux remote from the plume, and flame height.  In this section, the results from FDS simulations of the fire generated by the natural gas burner experiments are compared with the experimental results.  The average plume temperatures at elevations from 0.1~m through 1.2~m above the top face of the burner in 0.1~m increments are compared.

\subsection{Grid Size Analysis}
Figure~\ref{FHNG01_RI} shows the comparison of the centerline gas burner temperatures with three different sets of FDS results.  The only variable between these simulations was the RI. The RIs included three different computational cell sizes: 80~mm (RI=5), 40~mm~(RI=10), and 20~mm~(RI=20).  The associated 95\% confidence limits are also shown the temperature value for each height.  The FDS simulated temperatures based on the course cell size of 80~mm (RI=5) had the best agreement with the temperatures along centerline of the plume.  The two simulations with the higher RI values provided less agreement than the course mesh.

@@

This seemed counter intuitive to the results expected from an increased RI.  However, in each of the FDS models, the results closest to the burner seem to indicate that the model may not be ``burning'' the appropriate amount of fuel (fuel rich), which resulted in temperatures significantly lower than measured.  The effect of this fuel rich area above the burner results in poorer agreement with increasing RI.  It appears the RI=10 and RI=20 models follow the correct trend, but the combustion zone moved to a higher elevation relative to what was measured.  For example, the values at 0.5~m and 0.8~m above the burner, from the RI=10 and RI=20 cases, respectively, seemed to coincide with the measured value at 0.3~m above the burner.  Adjusting the FDS corresponding predictions from the RI=10 and RI=20 models down to the 0.3~m position appears to move the remaining temperatures above those points into alignment with the measurements along the centerline of the plume. FDS is a complex model, so this explanation may be only one possibility.

These simulation results are consistent with the comparisons made to the McCaffrey's plume correlation from his study, Purely Buoyant Diffusion Flames~\cite{McCaffrey:1979}, in the FDS Validation Guide~\cite{FDS_Validation_Guide}. McCaffrey used a natural gas burner 0.30~m square, which is similar in size to the burner used in this study.  McCaffrey noted that, ``visually non-axisymmetric behavior could be observed only very close to the burner surface''.  In the McCaffrey study, temperatures were measured with Type K thermocouples covered with a 1.0~mm stainless steel sheath. The peak heat release rate in McCaffery's plume study was 57.5~kW, which was also the closest to the 80~kW fires in the current study.  The comparison between the McCaffrey plume correlation and the FDS simulations at RI values of 5, 10, and 20 are shown in Figure~\ref{McCaffrey_Plume_comp}. Focusing on the data between the burner surface and 1~m above the burner surface, FDS under-predicted the temperature near the surface of the burner, and then over-predicted as the elevation increased to approximately 0.4~m and 1.3~m above the burner for the RI=10 and 20 cases, respectively. Once the temperature comparisons were above the continuous and intermittent flame regions, the agreement between the cases and the correlation began to improve.  Similar to the current study, increased RIs did not improve the simulation of the vertical centerline temperatures close to the burner.

\begin{figure}[h]
	\includegraphics[width=5in] {../Figures/McCaffrey_Plume_comp}  \\
	\caption[Comparison of vertical centerline temperatures above a 0.3~m natural gas burner from McCaffrey with FDS simulated gas temperatures.] {Comparison of vertical centerline temperatures above a 0.3~m natural gas burner from McCaffrey with FDS simulated gas temperatures generated with three different RIs (5, 10, and 20) from the FDS Validation Guide~\cite{FDS_Validation_Guide}.}
	\label{McCaffrey_Plume_comp}
\end{figure}

The comparisons above used the predicted gas temperature from the model. Another comparison can be made examining the THERMOCOUPLE results from FDS to see if the model of the thermocouple provides improved agreement with the temperatures measured via physical thermocouples.  This comparison is shown in Figure~\ref{FHNG01_RI_TC}.  Not surprisingly, the sub model of the thermocouple device shifted the FDS values slightly.  At temperatures greater than 700$^\circ$C, the THERMOCOUPLE temperatures tended to be lower than the gas temperatures, perhaps due to radiant heat transfer losses. At temperatures less than 700$^\circ$C, the `THERMOCOUPLE' temperatures tended to be a higher than the gas temperatures, perhaps due to gains from radiant heat transfer.  The best agreement along the vertical array of temperatures was provided by the RI=5 model.  Variables such as density and emissivity required for modeling the thermal response, and heat transfer to the thermocouple bead, may have generated the differences between the gas temperature and the measured temperature FDS values.

\begin{figure}[h]
	\includegraphics[width=5in]{../Figures/FHNG01_RI_TC} \\
	\caption[Comparison of vertical centerline THERMOCOUPLE temperatures above the natural gas burner with FDS simulated gas temperatures.]{Comparison of vertical centerline `THERMOCOUPLE' temperatures above the natural gas burner with FDS simulated gas temperatures generated with three different RIs (5, 10, and 20).}
	\label{FHNG01_RI_TC}
\end{figure}

@@

The heat flux data used for comparison was from the Chapter 2 source fire characterization experiments, the measured heat flux at 0.95~m from the centerline of the burner and 0.5~m above the top of the burner.  In Figure~\ref{FHNG01_HF_RI}, the averaged values from both heat flux gauges are plotted against the FDS calculated heat flux at three different RIs of 5, 10, and 20.  These FDS simulations used the default radiative fraction value of~0.2.  As with the temperature comparisons, the course grid with an RI=5 provided the best agreement to the measurements, but all the values were lower than the measured heat flux.

\begin{figure}
	\includegraphics[width=5in]{../Figures/FHNG01_HF_RI} \\
	\caption[Comparison of the average heat flux measured 0.95~m from the centerline and 0.5~m above the top of the natural gas burner with FDS simulated heat flux values.]{Comparison of the average heat flux measured 0.95~m from the centerline and 0.5~m above the top of the natural gas burner with FDS simulated heat flux values generated with three different RIs (5, 10, and 20).}
	\label{FHNG01_HF_RI}
\end{figure}

Reviewing the heat flux data documented in Chapter 2 from both the 0.95~m  and the 2.0~m from the vertical centerline of the burner positions provided average steady state heat flux values of approximately 2.0~kW/m$^2$ and 0.43~kW/m$^2$, respectively.  Applying the point source method discussed in Chapter 7, values for the radiative fraction were determined to be in the range of 0.25 to 0.28.  Values found in the literature for natural gas range from 0.15 to 0.25~\cite{Modak:1977,Hostikka:3}.   
Therefore it was of interest to conduct FDS simulations with radiative fractions ranging from 0.2 to 0.3.  This was based on the default value used in FDS and the upper bound of the literature values as this range covered the radiative fraction values estimated from the experiments.  The course grid, RI=5, was used for the FDS models.  The comparison of the results is shown in Figure~\ref{FHNG01_HF_RI5}.  The radiative fraction of 0.22 provided the best agreement with the measurement.

@@

One of the outputs that can be gotten from FDS is a heat release rate per unit length (HRRPUL) that has been used to determine the height where energy stops being released along the vertical centerline of the gas burner~\cite{McDermott:2011}.  Estimates of the flame height were made by estimating the height where 95 to 99~$\%$ of the total energy had been released. To obtain a value based on a measurement for comparison to the HRRPUL, the peak visible flame height range estimate was developed.  The peak visible flame height ranged from the location where the flame existed only 10\% of the time with the 95\% certainty range.  The Table~\ref{tab:FDS_FH} displays estimated values from the models using all three RI values as well as the 95~$\%$ certainty range for the peak visible flame height from Chapter 2.  The course and medium grids, RI=5 and R=10, were not significantly different from the estimations of the peak visible flame height 95~$\%$ certainty value range.

@@

One of the advantages of a computational fluid dynamics model such as FDS is the ability produce enough spatial data in three dimensions to allow the generation of visualizations.  The impact of the mesh resolution can be seen in the three images of the fire shown in Figure~\ref{FDS_Flame_Image_comp}.  The increased resolution leads to a flame structure that provides a better match to the look of the actual flame.  However, in this scenario, the increased resolution within the flow field did not result in improved numerical agreement with the measured values.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=1.7in]{../Figures/NG_80kW_GBWall_2D_RI=05_0601crop}
  \includegraphics[width=1.7in]{../Figures/NG_80kW_GBWall_2D_RI=10_0601crop}
  \includegraphics[width=1.7in]{../Figures/NG_80kW_GBWall_2D_RI=20_0601crop}\\
  \caption[Flame images generated with FDS and Smokeview]{Flame images generated with FDS and Smokeview. The left image has a grid size of 80~mm (RI=5), the center image has a grid size of 40~mm (RI=10), and the right image has a grid size of 20~mm (RI=20).}
  \label{FDS_Flame_Image_comp}
\end{figure}

@@

The comparisons of the temperature, heat flux, and flame heights for this specific scenario showed the best agreement with the 80~mm grid (RI=5) FDS simulations.  Even though the differences between the radiative fractions of 0.2 and 0.22 were small, for the remaining simulation cases, 0.22 were used.  While the best agreement for the fire was demonstrated with the RI=5 grid, the RI=10 grid was used in the subsequent instrumented wall and the compartment model sections to see if the increased resolution comes into play as the models got additional obstructions and surfaces.

@@

The next comparisons were made with the natural gas source fire at four different distances from an instrumented gypsum board wall.  Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610~m), one burner width (0.305~m), one half burner width (0.150~m), and then touching the wall (0.0~m).  The simulations replicated the same distances.  The simulation included a vertical gypsum board wall 1.2~m wide that extended from the floor of the model up to the top of the domain.  The material properties of the wall are based solely on the properties of gypsum cited in Chapter 4.  The values compared in the section include vertical centerline temperatures above the burner, wall temperatures, and heat flux values at the position of the wall surface.

@@

The computational domain of the model was the same as the volume considered for the source fire simulations, a cube 2.4~m on a side. The models were run with two different-sized computational grid cells, an 80~mm (RI=5) grid and a 40~mm (RI=10) grid.  The face of the wall is located 1.95~m from the front of the domain.  
The wall was modeled with material properties for gypsum wallboard that were documented in Chapter 4.  The input properties were thermal conductivity (0.15 W/m K), specific heat (1.00 kJ/kg K), and density (807 kg/m$^3$).  In addition, FDS includes a default value of 0.9 for emissivity, and this value was representative of the painted surface of the gypsum wallboard. As noted earlier in this study, gypsum wallboard undergoes chemical and physical changes as it is heated.  These changes include the loss of homogeneity of the gypsum material, a change in density, and, as the paper is charred or burned, potential change in emissivity.  The thermal conductivity and the specific heat of the gypsum wallboard varies on the order of 20~$\%$ over the temperature range from 20 to 500$^\circ$C~\cite{Gross:1985}.  While FDS can accommodate changes in thermo-conductivity and specific heat with respect to time, density, and emissivity must remain constant~\cite{FDS_Users_Guide}.  For the purposes of this study, the input values will remain constant throughout the course of the simulations.  The only variable will be the distance between the edge of the burner and the wall.

As before, the FDS simulations included devices used to simulate thermocouples and total heat flux gauges.  These devices were positioned in the same locations as the locations used in the physical experiments, above the centerline of the burner and on the wall as shown in Chapter 4, Figure~\ref{Instrumented_GB_Wall}. In addition, HRRPUL was again used to gain insight into the peak flame height simulated by FDS.

@@

 The temperatures measured by thermocouples positioned along the vertical centerline of the natural gas burner was compared with FDS generated temperatures from two simulations, one with RI-5 and one with RI=10.  The measurement points were at intervals of 0.2~m from 0.2~m to 2.2~m above the burner.  Figure~\ref{FDS_Plume_IWGB_comp} contains four graphs.  Each graph shows the same types of comparison, and the only difference is the distance away from the gypsum wallboard.  Beginning with the top left graph position and viewing left to right, the distances from the wall are 0.61~m, 0.31~m, 0.15~m, and 0~m respectively.
